---
title: "R Exercises"
author: |
  | Consulting for Statistics, Computing and Analytics Research

date: '`r Sys.Date()`'
output:
  html_notebook:
    theme: sandstone
    font-family: Roboto
    highlight: zenburn
    code_folding: hide
---
 
```{r setupExercises, include=FALSE}
knitr::opts_chunk$set(eval=F, cache = F, message = F, warning = F, R.options=list(width=120), comment=NA)
```


1. Create an object that consists of the numbers one through five using `c(1,2,3,4,5)` or `1:5`

```{r}
x = c(1,2,3,4,5)
x = 1:5
```


2. Create a different object, that is the same as that object, but plus 1 (i.e. the numbers two through six).

```{r}
z = x + 1
```


3. Without creating an object, use <span class="func">cbind</span> and <span class="func">rbind</span>, feeding the objects you just created as arguments. For example `cbind(obj1, obj2)`.

```{r}
new_data = cbind(x, z)
new_data = rbind(x, z)
```


4. Create a new object using <span class="objclass">data.frame</span> just as you did <span class="func">rbind</span> or <span class="func">cbind</span> in #3.

```{r}
new_data = data.frame(x, z)
```


5. Inspect the class of the object, and use the <span class="func">summary</span> function on it.

```{r}
class(new_data)
summary(new_data)
```
